### 2022-02-23 preregistrations/1_pilot/preregistration.Rmd
intermediate update to prereg study1

```

@@

We assessed state anxiety during the experiment using the state subscale of the State-Trait Anxiety Inventory [STAI-S; @spielberger_1999]. The STAI-S contains 20 short items measuring current anxiety (e.g., "I feel tense"; See [Table S1](#appendix) for an overview of all items). Participants rated each item on a scale of 1 (not at all) to 4 (very much so). An overall state anxiety variable was computed by averaging across the 20 unweighted items (*M* = `r #TBD `, *SD* = `r #TBD `, $\alpha$ = `r #TBD `).

@@

Violence exposure was measured using the Neighborhood Violence Scale [NVS; @frankenhuis_2018; @frankenhuis_deVries_2020] and two items assessing exposure to physical fights before age 13. The NVS contains seven items measuring perceived exposure to violence before age 13 (e.g., "Crime was common in the neighborhood where I grew up"; See [Table S2](#appendix) for an overview of all items). Participants rated each on a scale from 1 (never true) to 5 (very often true). The physical fighting items assessed the number of times participants witnessed fights before age 13: "Based on your experiences, how many times did you see or hear someone being beaten up in real life, before age 13?" and "How many times were you in a physical fight, before age 13?" Answers to both items ranged from 1 (0 times) to 8 (12 or more times). The items of the NVS were averaged together (*M* = `r #TBD `, *SD* = `r #TBD `, $\alpha$ = `r #TBD `). Similarly, we averaged the scores on the two fighting items together. Finally, we created a perceived violence exposure composite by standardizing the NVS and fighting composites and calculating an unweighted average (*r* = `r #TBD `).

@@

We computed several unweighted unpredictability variables to be able to assess how they would relate to the dependent variables as well as to each other: 1) overall perceived unpredictability variable containing all eight items of the perceived unpredictability scale (*M* = `r #TBD `, *SD* = = `r #TBD `, $\alpha$ = `r #TBD `); 2) an overall QUIC variable by averaging all 37 items (*M* = `r #TBD `, *SD* = = `r #TBD `, $\alpha$ = `r #TBD `); 3) five separate QUIC subscales by averaging their sub-items (Parental monitoring and involvement: *M* = `r #TBD `, *SD* = = `r #TBD `, $\alpha$ = `r #TBD `; Parental predictability: *M* = `r #TBD `, *SD* = = `r #TBD `, $\alpha$ = `r #TBD `; Parental environment: *M* = `r #TBD `, *SD* = = `r #TBD `, $\alpha$ = `r #TBD `; Physical environment: *M* = `r #TBD `, *SD* = = `r #TBD `, $\alpha$ = `r #TBD `; Safety and security: *M* = `r #TBD `, *SD* = = `r #TBD `, $\alpha$ = `r #TBD `). See also [below](#efa) for details on a factor analysis based on the individual items across scales.

One additional scales measured the stability of the family and social environment. On a scale of 1 (the same all the time) to 5 (constant and rapid changes), participants indicated how often the following aspects of their family and social environment changed before age 13: 1) economic status, 2) family environment, 3) childhood neighborhood environment, and 4) childhood school environment.

@@

We assessed household chaos before age 13 using an adaptation of the Confusion, Hubbub, and Order Scale [CHAOS; @matheny_1995]. The CHAOS consists of 15 items measuring the level of chaos in the household (e.g., "No matter how hard we tried, we always seemed to be running late"; See [Table S5](#appendix) for an overview of all items). We made two changes to the original scale as described in @matheny_1995. First, all items were converted from the present tense to the past tense, and were endorsed as applying to participants' lives before age 13. Second, all items were rated on a scale of 1 (never true) to 5 (very often true) instead of the original yes/no answer format. This change was also implemented to reduce cognitive load by keeping the answer options the same between scales. An overall household chaos variable was computed by averaging the 15 unweighted items (*M* = `r #TBD `, *SD* = `r #TBD `, $\alpha$ = `r #TBD `).

@@

Participants' perceived level of resource scarcity before age 13 was measured using seven items (e.g., "Your family had enough money to afford the kind of home you all needed"; See [Table S6](#appendix) for an overview of all items). Participants rated each item on a scale from 1 (never true) to 5 (very often true). Scores for the first six items were reverse coded so that higher scores indicated more perceived resource scarcity. The items were averaged together to create an unweighted composite scale (*M* = `r #TBD `, *SD* = `r #TBD `, $\alpha$ = `r #TBD `).

In addition, we measured several indicators of objective SES before age 13. First, participants separately indicated the highest education of their mother and father on an 8-point scale: 'some high school', 'GED', 'high school diploma', 'some college but no college degree', associate's  degree', 'bachelor's degree', 'master's  degree', or 'doctoral or lab degree'. The mother and father education level were averaged to create an overall unweighted parental education composite (*M* = `r #TBD `, *SD* = `r #TBD `). Participants also indicated their family's household income before age 13 on a 6-point scale: 'less than \$ 25k/year', '\$25k - \$49k/year, '\$50 - \$74k/year', '\$75 - \$99k/year', '\$100 - \$149k/year', 'more than \$150k/year'. Scores were reverse coded so that higher scores indicated higher levels of poverty.

@@

>>>>

load(here("preregistrations", "1_pilot", "staged_objects.Rmd"))
```

@@

We assessed state anxiety during the experiment using the state subscale of the State-Trait Anxiety Inventory [STAI-S; @spielberger_1999]. The STAI-S contains 20 short items measuring current anxiety (e.g., "I feel tense"; See [Table S1](#appendix) for an overview of all items). Participants rated each item on a scale of 1 (not at all) to 4 (very much so). An overall state anxiety variable was computed by averaging across the 20 unweighted items (*M* = `r txt_ivs$stai_s_mean$mean`, *SD* = `r txt_ivs$stai_s_mean$sd`, $\alpha$ = `r #TBD `).

@@

Violence exposure was measured using the Neighborhood Violence Scale [NVS; @frankenhuis_2018; @frankenhuis_deVries_2020] and two items assessing exposure to physical fights before age 13. The NVS contains seven items measuring perceived exposure to violence before age 13 (e.g., "Crime was common in the neighborhood where I grew up"; See [Table S2](#appendix) for an overview of all items). Participants rated each on a scale from 1 (never true) to 5 (very often true). The physical fighting items assessed the number of times participants witnessed fights before age 13: "Based on your experiences, how many times did you see or hear someone being beaten up in real life, before age 13?" and "How many times were you in a physical fight, before age 13?" Answers to both items ranged from 1 (0 times) to 8 (12 or more times). The items of the NVS were averaged together (*M* = `r txt_ivs$violence_mean$mean`, *SD* = `r txt_ivs$violence_mean$sd`, $\alpha$ = `r #TBD `). Similarly, we averaged the scores on the two fighting items together. Finally, we created a perceived violence exposure composite by standardizing the NVS and fighting composites and calculating an unweighted average (*r* = `r txt_cors$vio$r`).

@@

We computed several unweighted unpredictability variables to be able to assess how they would relate to the dependent variables as well as to each other: 1) overall perceived unpredictability variable containing all eight items of the perceived unpredictability scale (*M* = `r txt_ivs$unp_mean$mean`, *SD* = = `txt_ivs$unp_mean$sd`, $\alpha$ = `r #TBD `); 2) an overall QUIC variable by averaging all 37 items (*M* = `r txt_ivs$quic_total_mean$mean`, *SD* = = `r txt_ivs$quic_total_mean$sd`, $\alpha$ = `r #TBD `); 3) five separate QUIC subscales by averaging their sub-items (Parental monitoring and involvement: *M* = `r txt_ivs$quic_monitoring_mean$mean`, *SD* = `r txt_ivs$quic_monitoring_mean$sd`, $\alpha$ = `r #TBD `; Parental predictability: *M* = `r txt_ivs$quic_par_predict_mean$mean`, *SD* = `r txt_ivs$quic_par_predict_mean$sd`, $\alpha$ = `r #TBD `; Parental environment: *M* = `r txt_ivs$quic_par_env_mean$mean`, *SD* = `r txt_ivs$quic_par_env_mean$sd`, $\alpha$ = `r #TBD `; Physical environment: *M* = `r txt_ivs$quic_phys_env_mean$mean`, *SD* = `r txt_ivs$quic_phys_env_mean$sd`, $\alpha$ = `r #TBD `; Safety and security: *M* = `r txt_ivs$quic_safety_mean$mean`, *SD* = `r txt_ivs$quic_safety_mean$sd`, $\alpha$ = `r #TBD `). See also [below](#efa) for details on a factor analysis based on the individual items across scales.

One additional scale measured the stability of the family and social environment. On a scale of 1 (the same all the time) to 5 (constant and rapid changes), participants indicated how often the following aspects of their family and social environment changed before age 13: 1) economic status, 2) family environment, 3) childhood neighborhood environment, and 4) childhood school environment.

@@

We assessed household chaos before age 13 using an adaptation of the Confusion, Hubbub, and Order Scale [CHAOS; @matheny_1995]. The CHAOS consists of 15 items measuring the level of chaos in the household (e.g., "No matter how hard we tried, we always seemed to be running late"; See [Table S5](#appendix) for an overview of all items). We made two changes to the original scale as described in @matheny_1995. First, all items were converted from the present tense to the past tense, and were endorsed as applying to participants' lives before age 13. Second, all items were rated on a scale of 1 (never true) to 5 (very often true) instead of the original yes/no answer format. This change was also implemented to reduce cognitive load by keeping the answer options the same between scales. An overall household chaos variable was computed by averaging the 15 unweighted items (*M* = `r txt_ivs$chaos_mean$mean`, *SD* = `r txt_ivs$chaos_mean$sd`, $\alpha$ = `r #TBD `).

@@

Participants' perceived level of resource scarcity before age 13 was measured using seven items (e.g., "Your family had enough money to afford the kind of home you all needed"; See [Table S6](#appendix) for an overview of all items). Participants rated each item on a scale from 1 (never true) to 5 (very often true). Scores for the first six items were reverse coded so that higher scores indicated more perceived resource scarcity. The items were averaged together to create an unweighted composite scale (*M* = `r txt_ivs$ses_subj_mean$mean`, *SD* = `r txt_ivs$ses_subj_mean$sd`, $\alpha$ = `r #TBD `).

In addition, we measured several indicators of objective SES before age 13. First, participants separately indicated the highest education of their mother and father on an 8-point scale: 'some high school', 'GED', 'high school diploma', 'some college but no college degree', associate's  degree', 'bachelor's degree', 'master's  degree', or 'doctoral or lab degree'. The mother and father education level were averaged to create an overall unweighted parental education composite (*M* = `r 8-txt_ivs$edu_parents_recoded$mean`, *SD* = `r txt_ivs$edu_parents_recoded$sd`). Participants also indicated their family's household income before age 13 on a 6-point scale: 'less than \$ 25k/year', '\$25k - \$49k/year, '\$50 - \$74k/year', '\$75 - \$99k/year', '\$100 - \$149k/year', 'more than \$150k/year'. Scores were reverse coded so that higher scores indicated higher levels of poverty.

@@

# Deviations from preregistration

- Creation of poverty composite

# Results

## Exclusions

## Primary analyses

### Change Detection
Violence exposure did not have an effect on the speed of change detection (see Figure 1). However, participants with high violence exposure were less accurate ($\beta$~median~  = `r #TBD`, `r #TBD`% of *p*s < .05).

we found a negative main effect of violence exposure on the drift rate ($\beta$~median~  = `r #TBD`, `r #TBD`% of *p*s < .05), indicating that participants with more violence exposure showed slower information processing.

### Attention Cueing Task
We did not find a significant interaction effect between violence exposure and condition on raw Attention Cueing performance (see Figure 2. There was a significant main effect of violence exposure (($\beta$~median~  = `r #TBD`) under `r #TBD`% and `r #TBD`% of analytic decisions on raw RTs and log-transformed RTs, respectively (not shown in this figure), indicating that participants with high violence exposure generally performed slower on the task.

The DDM analysis uncovered a significant interaction effect between violence exposure and task condition on the drift rate ($\beta$~median~  = `r #TBD`, `r #TBD`% of *p*s < .05). Simple slopes analysis suggested that on cued trials, high violence exposure participants tended to have a lower drift rate than low violence exposure participants (`r #TBD`% ps < .05). This difference disappeared on neutral trials, with both groups showing comparable drift rates (0% ps < .05) (See Table 2).

We did not find significant interaction effects for boundary separation and non-decision time, although we did find significant main effects of violence exposure on boundary separation ($\beta$~median~  = `r #TBD`, `r #TBD`% of *p*s < .05) and non-decision time ($\beta$~median~  = `r #TBD`, `r #TBD`% of *p*s < .05), suggesting that participants with more violence exposure had more conservative decision boundaries and were slower to start the decision-making process (possibly reflecting slower eye-movements towards the targets).

### Flanker Task
We found a significant interaction effect between violence exposure and task condition on raw Flanker performance in `r #TBD`% and `r #TBD`% of analytic decisions for raw RTs and log-transformed RTs, respectively (see Figure 3). Simple slopes analyses showed that under `r #TBD`% of analytic decisions, participants with more violence exposure were slightly slower on congruent trials. This difference disappeared across all analytic decisions on incongruent trials.

The DDM analyses showed a combination of parameter differences that suggest a compensatory mechanism in people with more violence exposure. First, people with more violence exposure extracted lower-quality perceptual information (p) from the flankers ($\beta$~median~  = `r #TBD`, `r #TBD`% of *p*s < .05), which is analogous to the lower drift rates on the other tasks. Thus, they needed more time to gather enough information to reach the decision boundary. Second, the initial width of the attention spotlight (sda) did not differ across levels of violence exposure, which is not in line with our expectations of a broader, more diffuse attention style in people from adversity.
Third, people with more violence exposure had a higher shrinking rate rd ($\beta$~median~  = `r #TBD`, `r #TBD`% of *p*s < .05), meaning that they were faster at narrowing down their attention towards the central arrow (thereby decreasing flanker interference quicker, as shown by the sda/rd ratio). Judging from the interaction effect on raw RTs, this higher shrinking rate seems to offset the negative effect of lower perceptual input on incongruent trials. Thus, contrary to expectations, participants with more violence exposure seem to experience less interference from the flanker arrows.

## Secondary analyses

### Factor structure of unpredictability items

We conducted an EFA including all items measuring unpredictability: 1) The QUIC items, 2) the perceived unpredictability items, 3) the CHAOS items, 4) the items measuring environmental change, 5) number of partners of the father and mother, 6) Number of residential changes, 7) household size.

The measures of residential changes and number of different partners of both parents were heavily positively skewed.

Two participants indicated 50 different partners for either their mother or father. This number corresponded with other items focusing on the family situation, such as on the QUIC ("At least one of my parents had many romantic partners") and changes in the family environment. Therefore, we saw no reason to exclude these responses. We binned the responses for these three variables to get rid of the skew. For number of residential changes, responses were recoded into 7 bins: 0 (0 times), 1 (1-2 times), 2 (3-4 times), 3 (5-6 times), 4 (7-8 times), 5 (9-10 times), and 6 (10 or more times). For the number of partners of both parents, responses were recoded into .. bins: 0 (0 partners), 1 (1 partner), 2 (2 partners), 3 (3 partners), 4 (4 partners), 5 (5 partners), and 6 (6 or more partners).
